---
title: "Simulation Framework"
output: pdf_document
date: "May 2024"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

Here we describe the simulation framework used to test and analyse our modeling framework. We utilize functions from library `ape` and `phytools` to complete this simulation. 

The simulation framework begins with generating a phylogenetic tree. We provide code below. Here, we call the phylogenetic tree used for our simulation experiments. 

### 1. Simulate phylogeny

```{r echo = FALSE}
rm(list = ls())

# Some libs
library(ape)
library(phytools)

# Root directory
root <- "~/nicheEvol_ms"

# Source required functions
source(file.path(root, "scripts/00-functions.R"))

# Filepaths
data_directory <- file.path(root, "data")
raw_directory <- file.path(root, "raw_data")

# Config params
SIM <- TRUE
TREE_NAME <- "standard_sim"
TREE_TIME <- 10
MEAN_ROOT_VECTOR <- as.numeric(c(0, 0))
COVAR_TRAITS <- matrix(c(0.5, 1, 1, 2.5), nrow = 2, ncol = 2)
NTRAITS <- 2
SIGMA2 <- 1
ALPHA <- 1
USER_GRID <- FALSE
USER_PATH <- NULL
NSAMPLES <- 100
BINARY <- TRUE
EPATH <- "~/env"
```

```{r tree}
tree_time <- 8
tr <- ape::rbdtree(0.1, 0, Tmax = tree_time)
tr$edge.length <- tr$edge.length / max(phytools::nodeHeights(tr)[ ,2]) * 1
```

```{r pressure, echo=FALSE}
 # Load in object tree
contents <- load(file.path(root, "data/trees", paste0(TREE_NAME, ".Rdata")))
plot(tree)
```

### 2. Set up a spatial grid 

The spatial grid is set up systematically to sample points from the rasters below. 

```{r}
GRID <- createGrid(
  USER_GRID,
  nsamples = 100, USER_PATH,
  xmin = -10, xmax = 10,
  ymin = -10, ymax = 10
	)

env <- wrangleEnv(epath1 = "~/env", grid1 = GRID)
env_files <- file.path("~/env", c("CHELSA_bio_1.tif", "CHELSA_bio_13.tif"))
ras <- terra::rast(env_files)
extent = c(-81, -60, -40, 0)
ras <- terra::crop(ras, terra::ext(extent))
terra::plot(ras)
```


### 3. Simulate species-environment coefficients using an evolutionary process 

For a specific set of parameters and a given tree, the species-environment coefficients (niche parameters) are estimated using either the Ornstein-Uhlenbeck (OU) Process or the Brownian Motion (BM) Process. 

```{r, run = FALSE}
TREE <- tree
NTRAITS <- 2
SIGMA2 <- 1
ALPHA <- 1 
    if (ALPHA == 0) {
        bb <- phytools::fastBM(TREE, sig2 = SIGMA2, a0 = 1, nsim = NTRAITS)
    } else {
        bb <- phytools::fastBM(TREE,
            alpha = ALPHA, theta = 1,
            sigma2 = SIGMA2, nsim = NTRAITS
        )
    }
    colnames(bb) <- paste0("trait-", seq_len(ncol(bb)))

```

### 4. Simulate species' abundance or presence/absence 

Using the environmental variable, the spatial grid and the niche parameters, we can now simulate a species' abundance or presence/absence. We then split the species' data into train and test sets (where 70% of the data is training data and 30% is held-out for validation). Here, we show an example dataset for presence-absence data. 

```{r}
    # Simulates an OU/BM process
    all_data <- simulateAbundanceData_new(
        grid = GRID,
        nsamples = NSAMPLES,
        binary = BINARY,
        print = TRUE,
        tree,
        vars = c("CHELSA_bio_1.tif", "CHELSA_bio_13.tif"),
        ntraits = NTRAITS,
        sigma2 = SIGMA2,
        alpha = ALPHA,
        epath = EPATH
    )
```

```{r, echo = FALSE}
    Ytrue <- all_data$Ytrue
    Ytrue <- Ytrue[, -c(which(colnames(Ytrue) == "rwid"))]
    X <- all_data$X
    X <- X[, -c(which(colnames(X) == "rwid"))]

    # Subsetting matrices into test and train
    Ytrain <- all_data$Ytrain
    Ytrain <- Ytrain[, -c(which(colnames(Ytrain) %in% c("rwid", "tot")))]
    Y_occ <- Ytrain
    Y_occ[Y_occ < 0] <- 0
    Xtrain <- all_data$Xtrain
    Xtrain <- Xtrain[, -c(which(colnames(Xtrain) == "rwid"))]
    true_betas <- all_data$true_betas
    train_index <- all_data$train_index
    true_tree <- tree

    # Making matrices for stan code
    ns <- ncol(Ytrain)
    nsamples_train <- nrow(Ytrain)
    L <- matrix(ape::vcv(tree, corr = TRUE), ns, ns)
    distance_matrix <- ape::cophenetic.phylo(tree)
    I <- matrix(0, ns, ns)
    diag(I) <- 1

    I_y <- matrix(0, nsamples_train, nsamples_train)
    diag(I_y) <- 1

    standata <- list(
        N = nsamples_train,
        K = ncol(Xtrain),
        J = ns,
        E = 1,
        N_occ = nsamples_train,
        y = as.matrix(Ytrain),
        y_occ = as.matrix(Y_occ),
        x = Xtrain,
        x_occ = Xtrain,
        x_tilde = Xtrain,
        N_tilde = nrow(Xtrain),
        L_dist = distance_matrix,
        L_corr = L,
        beta_mu = rep(0, ncol(Ytrain)),
        true_betas = true_betas,
        I = I,
        Iy = I_y,
        alpha = 1,
        st_devs = rep(1, ncol(Ytrain)),
        U = 6,
        a = 3,
        b = 0.5,
        tr = tree,
        Y_true = Ytrue,
        X_true = X,
        train_index = train_index,
        rho_mean = 2.5,
        rho_sd = 0.2,
        alpha_mean = 0,
        alpha_sd = 0.2
    )

    # filename to write out
    filename <- paste0("example", ".Rdata")

    # Adding config params to the data file
    config <- list()
    config$sim <- SIM
    config$tree_name <- TREE_NAME
    config$tree_time <- TREE_TIME
    config$mean_root_vector <- MEAN_ROOT_VECTOR
    config$covar_traits <- COVAR_TRAITS
    config$ntraits <- NTRAITS
    config$sigma2 <- SIGMA2
    config$alpha <- ALPHA
    config$user_grid <- USER_GRID
    config$user_path <- USER_PATH
    config$nsamples <- NSAMPLES
    config$binary <- BINARY
    config$epath <- EPATH

    everything <- list(
        data = standata,
        config = config
    )

    # Save
    save(everything, file = paste0(
        data_directory, "/",
        filename
    ))
    
head(Ytrain)
```


This code and all required functions are available at the associated GitHub Repository. 
